Stage 1 Registered Report: The error when inferring phylogenies with incipient species by a Birth-Death model

### 2018-04-03 article.tex
No pilot study

The simulated alignment is the data that could be measured in the field.
This alignment is used to infer a posterior of, using BEAST2. We assume a
JC69 nucleotide substitution (as we do use that), a strict clock model (as
we have that) and a Birth-Death tree prior (as this simplification is the goal
of this research). Additionally, we assume a MRCA prior with a normal distribution
with a mean of the crown age, and a standard deviation of 0.001 time units (which
equates to 1000 years). After discarding a burn-in of 10\%, the MCMC algorithm must 
result in an ESS of the posterior of at least 200 and the MCMC chain length is 
chosen as such that this is the case for all simulations.

The similarity of each phylogeny in the posterior with 'the true species tree'
is determined with the nLTT statistic. In this way, a distribution of phylogenies
is transformed to a distribution of nLTT statistics.

To measure if sampling (either the youngest or oldest species-to-be to represent
a species) matters, for all parameter combinations that have incipient species,
we simulate 1000 incipient species trees that result in different species trees
when sampled either way. From each species tree, we compare the nLTT distributions
derived from the posterior. We quantify the Kullback-Leibler divergence (more
or less: the weighted fraction of overlapping nLTT 
values, which is zero if the two distributions are completely seperate, and close
to one if they overlap). For each parameter combination, we take the average
divergence. We assign verbal labels to this average divergence as such: 
0.0-0.25: clear effect, 0.25-0.5: strong effect, 0.5-0.75: mild effect, 
0.75-1.0: no effect. Per parameter setting, we show the average divergence and the 
verbal label. The distribution of these divergences is shown in the 
supplementary information.

To measure if protractedness matters, we compare the nLTT distribution per
combination of speciation initiation and extinction rate for a protracted
setting with its non-protracted (with a high speciation completion 
rate) setting. So, for each speciation initiation and extinction rate, 
the nLTT statistic distribution of the 
non-protracted setting is used as the baseline. That distribution is compared
with each protracted setting. We quantify the Kullback-Leibler divergence 
of these nLTT distributions as described above. Per protracted parameter setting, 
we generate 1000 replicate incipient species trees, each sampled randomly. We show the 
average divergence of these 1000 replicates and the verbal label.
The distribution of these divergences is shown in the supplementary information.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Pilot experiments}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Do the DNA sequence length and mutation rate (as described in the Methods)
give a good enough inference? Answer: pick the setting with the highest
expected number of taxa (highest speciation initiation and completion rate, 
no extinction). Simulate 1000 trees and obtain an nLTT distribution for a
reference DNA sequence length and mutation rate. Increasing DNA sequence length
or mutation rate increases the resolution of the experiment. This resolution
must not be too low (this would cost accuracy), nor too high (this would cost
too much computational resources). If doubling DNA length or DNA mutation rate
results in an nLTT distribution with 90% overlap with the reference distribution,
in 900 of 1000 cases, the (lower) reference resolution is considered high enough.

What is the MCMC chain length that gives a posterior ESS of 200 in all cases?
Answer: run all simulations, measure the posterior's ESS. If at least one 
ESS is lower than 200, elongate all MCMC chain lengths by a fraction 
to predict to increase it to 220 (for example, would the lowest ESS be 165, 
making the MCMC chains 33\% longer would bring it to (165\%*1.33=) 220. We use
220 to prevent an asymptotic approaching to 200.

>>>>

The simulated alignment is the data that could be measured in the field.
This alignment is used to infer a posterior of, using BEAST2. We assume a
JC69 nucleotide substitution (as we do use that), a strict clock model (as
we have that) and a Birth-Death tree prior (as this simplification is the goal
of this research). Additionally, we assume a MRCA prior with a normal distribution
with a mean of the crown age, and a standard deviation of 0.001 time units. 0.001
time units equates to 1000 years, which is the same resolution as which the
alignment is created. After discarding a burn-in of 10\%, the MCMC algorithm must 
result in an ESS of the posterior of at least 200 and the MCMC chain length is 
chosen as such that this is the case for all simulations. For simulations
with an ESS below 200, their MCMC chain length is elongated. An
MCMC chain is elongated by the proportion that is expected to increase it to 220. 
For example, would an ESS be 165, increasing its MCMC chain length by 33\% 
would bring it to (165\%*1.33=) 220. We use
220 as a safeguard to prevent an asymptotic approach to 200.

The nLTT statistic is used to quantify the similarity of each posterior's phylogeny 
with 'the true species tree'. In this way, a distribution of phylogenies
is transformed to a distribution of nLTT statistics. With this, the generative
model of this experiment has been completely described, from parameter estimates
to data (nLTT statistics) generated.

To quantify the effect of sampling (either the youngest or oldest species-to-be to represent
a species), for each parameter set that allows for incipient species,
we simulate 1000 incipient species trees that result in different species trees
when sampled either way. From each species tree, we compare the nLTT distributions
for the two sampling methods. These distributions are compared 
using the method described in [Kruschke 2013], with a non-informative
prior. In brief: if the 95\% Highest Density Interval of the differences between 
these two distributions excludes zero, it is credible the distributions are different.

To measure if protractedness matters, we compare the nLTT distribution per
combination of speciation initiation and extinction rate for a protracted
setting with its non-protracted (with a high speciation completion 
rate) setting. So, for each speciation initiation and extinction rate, 
the nLTT statistic distribution of the 
non-protracted setting is used as the baseline. That distribution is compared
with each protracted setting. This is done the same way as with the two
different sampling methods. [NOTE: we expect nLTT to go up for increasing SCR,
this should be in the method].

[NOTE: We do have more info, from Rampal's earlier articles. Perhaps this 
info can be put in]
